\chapter{Long Examples}
\RCSID$Id: ap04.tex,v 1.1 2002/08/23 14:58:45 nwalsh Exp $
\label{app:examples}

This appendix contains the source for several scripts that seemed
too long to place in the running text.  These scripts are provided
as examples only, they may not run on your system without some 
modification.

These examples are available online,
see the Preface for more information about retrieving online examples.

\bgroup
\makeatletter
\def\example@font{\tt\fontsize{8.42}{9.5pt}\selectfont}
\makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The \program{MakeTeXPK.pl} script builds PK files from \MF\ and PostScript
fonts.

\exampleinput{MakeTeXPK.pl}{ex:maketexpk}{MakeTeXPK.pl}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

The \program{MakeTeXTFM.pl} script builds TFM files from \MF\ and
PostScript (\ext{AFM} file) fonts.  

\exampleinput{MakeTeXTFM.pl}{ex:maketextfm}{MakeTeXTFM.pl}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

The \program{dvidxx.btm} script is written in \program{4DOS}'s extended
batch language.  It performs automatic font generation for \MF\
and PostScript fonts using \emTeX's drivers (version 1.4s).

\exampleinput{dvidxx.btm}{ex:dvidxx}{dvidxx.btm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

The \program{makepk.btm} script is used by \program{dvidxx.btm}, shown
in Example~\ref{ex:dvidxx} in this appendix.

\exampleinput{makepk.btm}{ex:makepk}{makepk.btm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

The \program{enc-afm.pl} script changes the encoding vector in an
\ext{AFM} file.

\exampleinput{enc-afm.pl}{ex:encafmpl}{enc-afm.pl}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

The \program{TeXtoXfont} script is a \Unix\ shell script (written in
\program{bash}) which builds an X11 font from a \MF\ or PostScript
font. 

\exampleinput{TeXtoXfont}{ex:textoxfont}{TeXtoXfont}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

The \program{PStoXfont} script is used by \program{TeXtoXfont}, shown
in Example~\ref{ex:textoxfont} in this appendix.  

\exampleinput{PStoXfont}{ex:pstoxfont}{PStoXfont}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

The \program{txt2verb.pl} script translates arbitrary text lines into
a format suitable for including in \verb|\verbatim| material in \TeX.

\exampleinput{txt2verb.pl}{ex:txt2verb}{txt2verb.pl}

\egroup